Remotely sensing river greenhouse gas exchange velocity using the SWOT satellite

### 2022-02-24 manuscript_v9.Rmd
prepping to rerun with wider k600

## Abstract
Extensive research over the past two decades has shown that the global river network emits significant amounts of greenhouse gas via gas exchange. Despite much progress, there is still large uncertainty in the temporal dynamics of gas exchange and thus carbon emissions to the atmosphere. Much of this uncertainty stems from uncertainty in gas exchange velocity $k_{600}$ (the rate of this diffusive transport). We propose that the NASA/CNES/USKA/CSA SWOT satellite, set to usher in a new age of hydrology research at the global scale, can provide similarly transformative insights to fluvial gas exchange modeling upon launch in 2022. Here, we present work inferring $k_{600}$ from synthetic future SWOT observations without in situ calibration. We exploit the unique geomorphology of SWOT-observable rivers to develop a physical model of gas exchange that is remotely sensible and explains `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2,2)*100`% of variation in $k_{600}$ We then couple this model with established Bayesian inference techniques to develop BIKER, the 'Bayesian Inference of the $k_{600}$ Exchange Rate' algorithm. We validate BIKER on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers and 166 discrete measurements of river gas exchange, yielding an algorithm that predicts the $k_{600}$ timeseries solely from SWOT observations with a by-river median Kling-Gupta Efficiency of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$KGE), 2)`. **get more impressive skill score than KGE** Similar to Bayesian remote sensing algorithms for SWOT, BIKER is better at reproducing the temporal dynamics of gas exchange (median correlation coefficient of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$r), 2)`), than the absolute rates of exchange (median relative bias of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$rBIAS), 2)*100`%). Finally, we show BIKER is robust to measurement errors implicit in the SWOT data. With SWOT set to launch in late 2022, we suggest that BIKER will be useful in mapping the global-scale spatiotemporal dynamics of fluvial gas exchange in large rivers.

@@

To answer this first question, we use hydraulic geometry- the fundamental geomorphic relationships between streamflow and channel shape [@leopoldHydraulicGeometryStream1953; @gleasonHydraulicGeometryNatural2015] to develop a process-based model for large-river (here defined as wider than 50m to align with SWOT) $k_{600}$. We then take these findings and explore the second question by implementing this hydraulic model, which in turn defines $k_{600}$, within an algorithm named BIKER ('Bayesian Inference of the $k_{600}$ Evasion Rate') to infer $k_{600$ solely from SWOT measurements. The goal of BIKER is to require no in situ inputs of any kind (although in situ data could be ingested and would improve results) such that it is globally implementable on any SWOT-observable river. We validate BIKER on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers (as SWOT has not yet launched) and explore BIKER's robustness to the expected measurement errors implicit in the satellite's observations. Finally, we also couple BIKER's $k_{600}$ estimates with $\delta CO_2$ to predict gas fluxes and compare these against established literature methods that are reliant on in situ hydraulic measurements.

@@

![Figure 2: Empirical testing of our large-river $k_{600}$ model on 166 measurements made in hydraulically-wide rivers. Dashed grey line denotes the 1:1 line. Note that axes are plotted in logarithmic space just for visualization: model fit and validation (via the coefficient of determination $r^2$) were calculated in linear space.](cache/k600_theory/k600Plot.jpg)

@@

- Correlation coefficient *r* to quantify accuracy of BIKER's temporal dynamics
- Root mean square error normalized by the observed mean (NRMSE) and prediction bias normalized by the observed mean (rBIAS) to assess bias
- Kling-Gupta Efficiency (KGE). KGE is frequently used to assess streamflow prediction and simultaneously assesses accuracy in both bias and dynamics. While a value greater than -0.41 means the model outperforms a uniform prediction of the mean [@knobenTechnicalNoteInherent2019], generally KGE scores are interpreted as being meaningful in ungauged settings if > 0.

@@

In that context, we sought to compare BIKER against established methods from the literature for upscaling carbon emissions from river networks (in this case, our `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])`  rivers) So, we pair the 26 biweekly $CO_2$ and water temperature samples from @beaulieuControlsGasTransfer2012 (section 3.1, figure S2) with every 14th set of daily SWOT observations (as the $CO_2$ data is bi-weekly). We then calculate $FCO_2$ using equation 1, an atmospheric $CO_{2_{air}}$ of 390 uatm, and a temperature dependent *Sc* was estimated following Raymond et al. (2012). Finally, we estimate a yearly total carbon emission rate (via $CO_2$ evasion) by applying each river’s mean $FCO_2$ over the river’s surface area and summing all rates across rivers.

@@

![Figure 4. Representative river timeseries plots of $k_{600}$. Orange is observed, while green is BIKER and uses SWOT measurements as its sole input. The green ribbon indicates the 95% CIs for the BIKER posteriors. Rivers are randomly sampled from the three tertiles of KGE scores. (a-b) are poorest performing rivers, (b-c) are in the middle, and (e-f) are the best performing rivers. Y axis is normalized by maximum observed values to compare visually.](cache/validation/timeseries_noerr_short.jpg)

Next, we calculate performance metrics following section 3.2.2 and Table S2. These are presented in Figure 5, which plots the scores for the 47 rivers with no measurement errors as empirical cumulative density functions (eCDFs). Median river $KGE$ is `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$KGE, na.rm=T),2)` and median river *r* is `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$r, na.rm=T),2)`. This indicates very strong inference of each river's temporal $k_{600}$ dynamics given that absolutely no in situ information is being used to predict $k_{600}$. NRMSE has a median score of `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$NRMSE, na.rm=T),2)` and median rBIAS is `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$rBIAS, na.rm=T),2)`, highlighting that many rivers which have significant positive biases (Figure 4 also confirms this visually). Taken in aggregate, Figures 4-5 indicate that BIKER is quite good at capturing temporal dynamics in $k_600$, however there is often positive bias in its estimates.

![Figure 5. Performance metrics by river, ploted as empirical cummulative density functions (eCDFs). Each subpanel is labelled by its performance metric (defined in Table S2). Dashed lines denote median scores.](cache/validation/validation_by_river.jpg)

@@

![Figure 6: Comparison of BIKER performance when considering measurement error in the SWOT observations for 16 rivers. Each subpanel is labelled by its performance metric (Table S2). Rivers that fall within the purple zone get worse when accounting for measurement error, while rivers in the green get better. Note also that axes are flipped in order to visualize all 'better performances' in the upper-right-corner of each sub-plot. Black line denotes linear regression while dashed grey line is the 1:1 line.](cache/validation/errorPlot.jpg)

**something something temporal variability**
You need to flesh this out a bit better and stop leaning so heavily on temporality. Feels almost like we're hiding something, which we aren't Colin was not a fan of the prior vs posterior CV eCDFs and also suggested that this should still hold even if we are doing just magntiude.

Given the promising *r* scores across all 47 rivers in Figure 5, we further explore BIKER's ability to infer temporal trends in $k_{600}$. Figure 7a plots a histogram of the slopes of linear regressions between BIKER-predicted $k_{600}$ and observed $k_{600}$. A slope of 1 indicates that BIKER correctly infers the daily flucuations in gas exchange (even if there is bias implicit in these estimates), while a slope well above or below 1 indicates incorrect inference of the temporal dynamics of $k_{600}$. Figure 6a shows that the majority of rivers' slopes approximate 1 (median slope: `r round(median(results_dynamics$Slope),2)`), indicating strong inference of temporal gas exchange trends.

![Figure 7: a) Histogram of the slopes of linear regressions between BIKER estimates and observed $k_{600}$. Slopes equaling 1 indicate correct inference of the temporal trends in daily $k_{600}$ (regardless of bias in the estimates). b) Empirical cummulative density distributions of the coefficient of varation of prior, posterior, and observed $k_{600}$ per river. If BIKER is correctly inferring temporal variation in $k_{600}$, the posterior eCDF will approximate the observed eCDF (as it does).](cache/validation/validation_temporal.jpg)

Finally, we explore whether BIKER improves upon the prior with respect to temporal variation in $k_{600}$, here defined using the coefficient of variation ($CV=\frac{SD}{mean}$). This is an essential test of a Bayesian method to determine if SWOT observations have improved our knowledge or if our initial guess proved statistically most useful. Figure 6b plots eCDFs of observed, prior, and posterior $k_{600}$ *CV*, strongly showing that BIKER improves upon the prior information available to us about $k_{600}$. Posterior $k_{600}$ *CV* approximates the observed eCDF and sufficiently 'closes the gap' between our prior knowledge on $k_{600}$ temporal variation and the actually observed variation in $k_{600}$. It must be stressed that this validation was performed using a relatively uninformed prior on $k_{600}$ (Text S3) and that this gap may not be as large in many settings with better prior knowledge. However, Figure 6b confirms that Bayesian techniques can be used to infer temporal variation in $k_{600}$ when good prior or in situ data are not available.

@@

![Figure 8: Yearly carbon emissions rate via $CO_2$ evasion across all rivers. Completely remotely-sensed methods are colored in red, in situ methods in purple, and the observed in orange.](cache/FCO2/FCO2_models.jpg)

>>>>

## Abstract
Extensive research over the past two decades has shown that the global river network emits significant amounts of greenhouse gas via gas exchange. Despite much progress, there is still large uncertainty in the temporal dynamics of gas exchange and thus carbon emissions to the atmosphere. Much of this uncertainty stems from uncertainty in gas exchange velocity $k_{600}$ (the rate of this diffusive transport). We propose that the NASA/CNES/USKA/CSA SWOT satellite, set to usher in a new age of hydrology research at the global scale, can provide similarly transformative insights to fluvial gas exchange modeling upon launch in 2022. Here, we present work inferring $k_{600}$ from synthetic future SWOT observations without in situ calibration. We exploit the unique geomorphology of SWOT-observable rivers to develop a physical model of gas exchange that is remotely sensible and explains `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2,2)*100`% of variation in $k_{600}$ We then couple this model with established Bayesian inference techniques to develop BIKER, the 'Bayesian Inference of the $k_{600}$ Exchange Rate' algorithm. We validate BIKER on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers and 166 discrete measurements of river gas exchange, yielding an algorithm that predicts the $k_{600}$ timeseries solely from SWOT observations with a by-river median Kling-Gupta Efficiency of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$KGE), 2)`. Similar to Bayesian remote sensing algorithms for SWOT, BIKER is better at reproducing the temporal dynamics of gas exchange (median correlation coefficient of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$r), 2)`), than the absolute rates of exchange (median relative bias of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$MAE), 2)*100`%). Finally, we show BIKER is robust to measurement errors implicit in the SWOT data. With SWOT set to launch in late 2022, we suggest that BIKER will be useful in mapping the global-scale spatiotemporal dynamics of fluvial gas exchange in large rivers.

@@

To answer this first question, we use hydraulic geometry- the fundamental geomorphic relationships between streamflow and channel shape [@leopoldHydraulicGeometryStream1953; @gleasonHydraulicGeometryNatural2015] to develop a process-based model for large-river (here defined as wider than 50m to align with SWOT) $k_{600}$. We then take these findings and explore the second question by implementing this hydraulic model, which in turn defines $k_{600}$, within an algorithm named BIKER ('Bayesian Inference of the $k_{600}$ Evasion Rate') to infer $k_{600}$ solely from SWOT measurements. The goal of BIKER is to require no in situ inputs of any kind (although in situ data could be ingested and would improve results) such that it is globally implementable on any SWOT-observable river. We validate BIKER on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers (as SWOT has not yet launched) and explore BIKER's robustness to the expected measurement errors implicit in the satellite's observations. Finally, we also couple BIKER's $k_{600}$ estimates with $\delta CO_2$ to predict gas fluxes and compare these against established literature methods that are reliant on in situ hydraulic measurements.

@@

![Figure 2: Empirical testing of our large-river $k_{600}$ model on 166 measurements made in hydraulically-wide rivers. Dashed grey line denotes the 1:1 line. Note that axes are plotted in logarithmic space just for visualization: model fit and validation (via the coefficient of determination $r^2$) were calculated in linear space.](cache/k600_theory/fig2.jpg)

@@

- Correlation coefficient *r* to quantify accuracy of BIKER's temporal dynamics
- Root mean square error normalized by the observed mean (NRMSE) to assess bias
- mean absolute residual error (MAE) to assess bias
- Kling-Gupta Efficiency (KGE). KGE is frequently used to assess streamflow prediction and simultaneously assesses accuracy in both bias and dynamics. While a value greater than -0.41 means the model outperforms a uniform prediction of the mean [@knobenTechnicalNoteInherent2019], generally KGE scores are interpreted as being meaningful in ungauged settings if > 0.

@@

In that context, we sought to compare BIKER against established methods from the literature for upscaling carbon emissions from river networks (in this case, our `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])`  rivers) So, we pair the 26 biweekly $CO_2$ and water temperature samples from @beaulieuControlsGasTransfer2012 (section 3.1, figure S2) with every 14th set of daily SWOT observations (as the $CO_2$ data is bi-weekly). We then calculate $FCO_2$ using equation 1, an atmospheric $CO_{2_{air}}$ of 400 uatm, and a temperature dependent *Sc* was estimated following Raymond et al. (2012). Finally, we estimate a yearly total carbon emission rate (via $CO_2$ evasion) by applying each river’s mean $FCO_2$ over the river’s surface area and summing all rates across rivers.

@@

![Figure 4. Representative river timeseries plots of $k_{600}$. Orange is observed, while green is BIKER and uses SWOT measurements as its sole input. The green ribbon indicates the 95% CIs for the BIKER posteriors. Rivers are randomly sampled from the three tertiles of KGE scores. (a-b) are poorest performing rivers, (b-c) are in the middle, and (e-f) are the best performing rivers. Y axis is normalized by maximum observed values to compare visually.](cache/validation/fig4.jpg)

Next, we calculate performance metrics following section 3.2.2 and Table S2. These are presented in Figure 5, which plots the scores for the 47 rivers with no measurement errors as empirical cumulative density functions (eCDFs). Median river $KGE$ is `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$KGE, na.rm=T),2)` and median river *r* is `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$r, na.rm=T),2)`. This indicates very strong inference of each river's temporal $k_{600}$ dynamics given that absolutely no in situ information is being used to predict $k_{600}$. NRMSE has a median score of `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$NRMSE, na.rm=T),2)`, highlighting that many rivers which have significant positive biases (Figure 4 also confirms this visually). Median MAE, however, is only `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$MAE, na.rm=T),2)` m/day. Taken in aggregate, Figures 4-5 indicate that BIKER is quite good at capturing temporal dynamics in $k_600$, however there is often positive bias in its estimates.

![Figure 5. Performance metrics by river, ploted as empirical cummulative density functions (eCDFs). Each subpanel is labelled by its performance metric (defined in Table S2). Dashed lines denote median scores. Note the log10 scale for MAE.](cache/validation/fig5.jpg)

@@

![Figure 6: Comparison of BIKER performance when considering measurement error in the SWOT observations for 16 rivers. Each subpanel is labelled by its performance metric (Table S2). Rivers that fall within the purple zone get worse when accounting for measurement error, while rivers in the green get better. Note also that axes are flipped in order to visualize all 'better performances' in the upper-right-corner of each sub-plot. Black line denotes linear regression to aid in visualization.](cache/validation/fig6.jpg)

Finally, we sought to further explore both BIKER's ability to infer temporal trends in $k_{600}$, as well as the influence of prior error on $k_{600}$.

Figure 7a plots a histogram of the slopes of linear regressions between BIKER-predicted $k_{600}$ and observed $k_{600}$. A slope of 1 indicates that BIKER correctly infers the daily fluctuations in gas exchange (even if there is bias implicit in these estimates), while a slope well above or below 1 indicates incorrect inference of the temporal dynamics of $k_{600}$. Figure 6a shows that the majority of rivers' slopes approximate 1 (median slope: `r round(median(results_dynamics$Slope),2)`), indicating strong inference of temporal gas exchange trends.

![Figure 7: a) Histogram of the slopes of linear regressions between BIKER estimates and observed $k_{600}$. Slopes equaling 1 indicate correct inference of the temporal trends in daily $k_{600}$ (regardless of bias in the estimates). b) Empirical cummulative density distributions of prior and posterior mean absolute error (MAE).](cache/validation/fig7.jpg)

Figure 7b comes prior error for $k_{600}$ against posterior error in $k_{600}$. We used the mean absolute error (Table S2), which is a metric that captures the average residual error in predictions, to assess this and plotted the results as eCDFS across all `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` rivers in the 'no-error' scenario. In Figure 7b, we see that BIKER reduces MAE, when the prior estimate was already quite good (approximately less than 3 m/day), while BIKER increases MAE if the prior estimate was already quite poor (again, approximately > 3 m/day). This indicates that our prior estimate of $k_{600}$ is too certain, i.e. BIKER is convinveed that the prior estimate is reasonable and, when combined with the hydraulics inference also in the model, yields a net worse performance. TO perturb this, we reran our BIKER validation using a less certain $k_{600}$ prior. SHOULD THIS BE SI OR JUST BE THE ACTUAL STUDY

@@

![Figure 8: Yearly carbon emissions rate via $CO_2$ evasion across all rivers. Completely remotely-sensed methods are colored in red, in situ methods in purple, and the observed in orange. Note that confidence intervals were not calculated for the remaining two models because their uncertainties are not available.](cache/FCO2/fig8.jpg)
